---
title: "Tutorial 1"
format: html
editor: visual
execute-dir: project
---

##### This tutorial introduces the structure, content, and usage of PLINK binary genotype datasets with 4,000 individuals and more than 300K SNPs. The goal is to help beginners understand how genotype data is stored and accessed in PLINK, and prepare them for downstream tasks like GWAS and QC.

You will need to install PLINK and run the analyses. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). PLINK is a free, open-source whole genome association analysis toolset, designed to perform a range of basic, large-scale analyses in a computationally efficient manner. For this tutorial, we are using simulated case/control data.

### 1. What is a PLINK Binary Dataset?

A PLINK binary dataset consists of three files:

-   **gwa.qc.bed** - Binary genotype matrix (compressed, encoded format). The reason why have this type of format is because the genotype matrix is very large, can have billions of entries, and have thousands of samples. So it is high dimensional. For that reason the file is compressed and it is encoded known as binary genotype file.
-   **gwa.qc.bim** - SNP annotation file: chromosome, SNP ID, base-pair position, alleles. It shows information of SNP stored in genotype matrix
-   **gwa.qc.fam** - Sample information file: IDs, sex, phenotype

Please download the three files from the following Google Drive link and place them in `02_activities/data/`: <https://drive.google.com/drive/folders/1v-FbDqu95wdUje3foFXMdXrBjYEAJTB2?usp=drive_link>

```{r setup, include=FALSE}
# Set up a consistent path for all chunks of code
knitr::opts_knit$set(root.dir = normalizePath("../../"))
 #This creates a consistent path for all the files.
```

```{r, message=FALSE, warning=FALSE, results='hide'}
library(data.table)
library(ggplot2)
library(seqminer)
library(HardyWeinberg)
library(dplyr)
```

### 2. View the .fam and .bim Files

Sample information file: IDs, sex, phenotype

```{bash}
pwd
head ./02_activities/data/gwa.qc.fam
```

| **Column Name** | **Description** |
|------------------------------------|------------------------------------|
| **FID** | Family ID (set to 0 if not used) |
| **IID** | Individual ID (unique for each person) |
| **PAT** | Paternal ID (set to 0 if unknown) |
| **MAT** | Maternal ID (set to 0 if unknown) |
| **SEX** | Sex: `1 = male`, `2 = female`, `0 = unknown` |
| **PHENOTYPE** | Trait status: `1 = control`, `2 = case`, `0` or `-9` = missing , or a quantitative trait |

```{bash}
# SNP annotation file: chromosome, SNP ID, base-pair position, alleles 
head ./02_activities/data/gwa.qc.bim
```

| **Column Name** | **Description** |
|------------------------------------|------------------------------------|
| **CHR** | Chromosome number |
| **SNP** | SNP ID (e.g., rs number) |
| **CM** | Genetic distance (usually 0) |
| **BP** | Physical position in base pairs, refers to specific position in genome |
| **A1** | PLINK's alternative allele |
| **A2** | the other allele (reference allele; mostly major alleles) |

Since we have 2 combinations for each chromosome in total 4 bases from 2 base pairs. But in these files we only represent the compressed version, only representing one of the bases in the pair.

### 3. Understanding the .bed file

The .bed file stores encoded genotypes in a compact binary format. It is not human-readable. Use PLINK to decode it.

Convert to readable format:

```{bash,results='hide',message=FALSE, warning=FALSE}
#plink2 --bfile ./02_activities/data/gwa.qc --export A --out ./02_activities/data/gwa_qc_additive
#Here is one example of additive coding. The export Option A is for additive effects

plink2 --bfile ./02_activities/data/gwa.qc --export A --out ./02_activities/data/gwa_qc_additive
```

This creates a .raw file with dosage coding:

0 = homozygous for A1

1 = heterozygous

2 = homozygous for A2 (REF alleles are counted by default). So basically we are counting the major allele as the reference.

It is very common to have missing data, but can add commands to remove remove these sites.

In the genotype matrix, we see that the rows represent different individuals, and for each we can see the identity of each SNP.

```{r}
# Load genotype data

geno <- fread("./02_activities/data/gwa_qc_additive.raw")  # columns include FID, IID, and SNP genotype
geno[1:5,1:10]
```

------------------------------------------------------------------------

### 4. Hardy-Weinberg Equilibrium (HWE) test

We can test each SNP for deviation from Hardy-Weinberg Equilibrium:

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc --hardy --out ./02_activities/data/gwa_qc_hwe
```

This produces gwa_qc_hwe.hardy containing observed and expected genotype counts plus p-values.

```{r,message=FALSE, warning=FALSE}
# View HWE results
hwe <- fread("./02_activities/data/gwa_qc_hwe.hardy")
head(hwe)
```

CHROM - Chromosome number.

ID - SNP ID (e.g. rsID).

A1, AX - The two alleles (effect and reference).

HOM_A1_CT - Count of A1/A1 homozygotes.

HET_A1_CT - Count of A1/AX heterozygotes.

TWO_AX_CT - Count of AX/AX homozygotes.

O(HET_A1) - Observed heterozygote frequency (proportion).

E(HET_A1) - Expected heterozygote frequency under HWE.

P - Hardy-Weinberg p-value testing deviation from equilibrium. This is the most important column

For QC, SNPs with very low P are often removed, as this may indicate genotyping error.

### 5. Estimate Allele Frequencies

Calculate and output allele frequencies in your dataset:

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc --freq --out ./02_activities/data/gwa_qc_freq
```

View frequency table: gwa_qc_freq.afreq includes columns for SNP, chromosome, allele counts, MAF.

```{r,message=FALSE, warning=FALSE}
freq <- fread("./02_activities/data/gwa_qc_freq.afreq")
head(freq)
```

Here the allele frequency counts the allele frequency that we are trying to estimate. It tells us what combination of alleles are counting.

### 6. Subset SNPs and Individuals

#### Step 1: Create SNP and sample list files

```{bash,message=FALSE, warning=FALSE}
# Create SNP list file with 4 example SNPs
setopt interactivecomments
#Here print some interesting SNPs that are printed to the file
printf "%s\n" rs1525873 rs2902663 rs343376 rs3813199 > ./02_activities/data/snplist.txt

# Create sample list file with header + 3 samples
#echo "FID IID" > ./02_activities/data/samplist.txt
#echo "0 A2001" >> ./02_activities/data/samplist.txt
#echo "2 A2003" >> ./02_activities/data/samplist.txt
#echo "5 A2006" >> ./02_activities/data/samplist.txt

```

#### Step 2: Subset SNPs and/or samples

```{bash, results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc \
      --extract ./02_activities/data/snplist.txt \
      --recode A \
      --out ./02_activities/data/geno.additive
#The bfile specific that path to file that need the analysis, it requires 3 files, if one of them is lost then, this will not run. 

# --keep ./02_activities/data/samplist.txt can be used to extract a specific list of samples
```

```{r}
# Load genotype data into R and make some operations.
geno_subset <- fread("./02_activities/data/geno.additive.raw")
#This will omit any data that has NA.
geno_subset=na.omit(geno_subset)
#In this case, code the phenotype column as a binary trait, that will be used in the logistic regression analysis, since the convention is that disease is coded as 1 and controls as 0, need to substract the the phenotype -1 since the original data the phenotype was coded as 2 and 1. 
geno_subset$PHENOTYPE=geno_subset$PHENOTYPE-1
head(geno_subset)
table(geno_subset$PHENOTYPE)
```

##### Alternatively, we can load all three files into R.

R can do the analysis, but I will be slower than PLINK2 since the program was designed to optimize the extraction process.

The PLINK2 commands are especially useful when working with a large number of SNPs (genome-wide), while R is better suited for loading smaller subsets of genotype data.

```{r,message=FALSE, warning=FALSE}
set.seed(1234)
fam_all <- fread("./02_activities/data/gwa.qc.fam")
N=dim(fam_all)[1];N
bim_all<- fread("./02_activities/data/gwa.qc.bim")
M <- nrow(bim_all);M

indx.SNP=which((bim_all$V2)%in%c('rs1525873','rs2902663','rs343376','rs3813199'))
#print(indx.SNP)
geno_mat<- readPlinkToMatrixByIndex('./02_activities/data/gwa.qc', 1:N, indx.SNP)
dim(geno_mat)
head(geno_mat)
```

Sometimes we not only need to extract a few samples but need to extract a massive number of samples

##### R functions to calculate the allele frequency.

```{r,message=FALSE, warning=FALSE}
geno_mat[geno_mat==-9]=NA
# Compute allele frequencies for each SNP (column), ignoring NAs
allele_freq <- apply(geno_mat, 2, function(x) {
  sum(x, na.rm = TRUE) / (2 * sum(!is.na(x)))
})

# Look at the result
allele_freq

```

##### We can also conduct the HWE test in R.

```{r,message=FALSE, warning=FALSE}

hwe_pvalue <- function(geno_vector) {
  # Remove missing
  geno_vector <- geno_vector[!is.na(geno_vector)]
  # Count genotypes
  counts <- table(geno_vector)
  
  # Check if there are enough counts to test
  if (any(counts == 0)) {
    return(NA)  # Cannot test properly with empty cell
  } else {
    result=HWExact(as.numeric(counts))$pval
    #result=HWChisq(as.vector(counts))$pval
    return(result)
  }
}
# Apply to all 4 SNPs (columns of the matrix)
hwe_pvalues <- apply(geno_mat, 2, hwe_pvalue)

# Show results
hwe_pvalues

#Compare with the HWE p-values produced by PLINK
hwe[hwe$ID%in%c('rs1525873','rs2902663','rs343376','rs3813199'),]
```

### 7. Logistic Regression: Additive Model

We talked that for Binary phenotype, when have affected and unaffected status, in that case, we need to perform a logistic regression model. In regression the probability of Y is linked to a variable and determine if there is a relationship between X and Y. In the case of logistic regression we are regressing a binary variable into a predictive variable. The goal of logisit c regresison is to model the prob of Y=1 or Y=0 using the predictor variable. Here logistic regression defins the prob of Y=1, so in this case need to do the log transformatio, of the model P(Y=1\|X)/1-P(Y=1\|X). We estiamte the beta1 since it modifies the probability of developing a particual disease, and the model will determine how significant is the influence of b1 in the prob of disease. Bae on the effect provided by th software can calcualte the probability of the disease by using the regression model.

```{r,message=FALSE, warning=FALSE}
#Pick SNP rs1525873. Here extract a genotype at specific genomic location
table(geno_subset$rs1525873_T)
freq[freq$ID=='rs1525873',] #freq is the file that load before in R and contains the allele frequency of SNPs in the population
```

The T allele is the major allele, with frequency 1 - 0.247497. Therefore, most samples have genotype coded as 2 rather than 0. When we look at the genotype matrix we do not see the nomenclature as A/a but obtain the actual base that represent the minor allele and the major allele.

```{r,message=FALSE, warning=FALSE}
# Fit logistic regression model
model <- glm(PHENOTYPE ~ rs1525873_T, data = geno_subset, family = binomial)

# View model summary
summary(model)
#I the model summary obtain the Estimate which represents the beta coeficient for the particular variant. 
table(geno_subset$rs1525873_T)
```

##### We visualize the association between the phenotype and the genotype.

```{r,message=FALSE, warning=FALSE}
# Make prediction over continuous range (for smooth curve)
# Create a new data frame containing a sequence of genotype values
 geno_range <- data.frame(
   rs1525873_T = seq(0, 2, length.out = 100)
 )
# Use the fitted logistic regression model to predict phenotype probability
# for each genotype value in geno_range
 geno_range$predicted_prob <- predict(model, newdata = geno_range, type = "response")

# Start from the observed genotype dataset
count_data <- geno_subset %>%
# Round phenotype values to the nearest integer (for binary phenotype: 0 or 1)
  mutate(PHENOTYPE = round(PHENOTYPE)) %>%  
# Remove rows with missing genotype or missing phenotype
  filter(!is.na(rs1525873_T), !is.na(PHENOTYPE)) %>%
 # Group data by genotype value and phenotype category 
  group_by(rs1525873_T, PHENOTYPE) %>%
# Count how many observations fall into each genotype-phenotype combination
  summarise(count = n(), .groups = "drop")

# Start building the plot
p1=ggplot() +
# Add points showing observed genotype-phenotype combinations
  geom_point(
    # Use the summarized count_data for the points
    data = count_data,
    # Map genotype to x-axis, phenotype to y-axis, and count to point size
    aes(x = rs1525873_T, y = PHENOTYPE, size = count),
    color = "blue"
  ) +
  # Add a regression line showing predicted probability across genotype values
  geom_line(# Use the prediction data frame for the line
    data = geno_range,
    aes(x = rs1525873_T, y = predicted_prob),
    color = "red",
    # Set line thickness
    size = 1.2
  ) +
  # Control the size range of the points
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Logistic Regression: PHENOTYPE ~ Genotype",
    x = "Genotype (0/1/2)",
    y = "Phenotype ",
    size = "Count"
  ) +
  # Restrict the visible x- and y-axis range
  coord_cartesian(xlim = c(-0.5, 2.5), ylim = c(-0.1, 1.1)) +
   # Apply a minimal theme for a clean appearance
  theme_minimal()

print(p1)
ggsave("./02_activities/data/p1_plot.png", plot = p1, width = 7, height = 5, dpi = 300)

```

In the Plot, can see the logistic regression model that was generated and for each genotype have 3 possibilities for the genotype on an additive model. Can see that individuals are distributed into 3 genotypes determined by the number of alleles homozygous WT, hetero and homozygous mutation. can see by the size of circles represent the amount of individuals in each genotype and the number of individual with genotype 2 is bigger in the disease status.

**Interpretation**: We performed logistic regression to test the association between genotype (rs1525873) and the binary phenotype. The regression coefficient for rs1525873 is 1.93140 (p \<2e-16), indicating that each additional copy of the T allele is associated with increased odds of being a case. The significant p-value suggests evidence of an association between this SNP and the phenotype in our sample.

In practice, when testing genetic associations, we include additional covariates in the model, such as age, sex, and principal components to adjust for confounding effects. We will cover this in more detail later in the course.

### 8. Logistic Regression: Dominance and Recessive Models

##### Dominance coding

```{r}
# Create a new data frame geno_subset_dom as a copy of geno_subset
geno_subset_dom <- geno_subset  

# Convert genotype values in columns 7 to the last column into dominant coding
geno_subset_dom[, 7:ncol(geno_subset_dom)] <- as.data.frame(
  lapply(geno_subset[, 7:ncol(geno_subset)], function(x) ifelse(x > 0, 1, 0))
)
head(geno_subset_dom)

```

Fit logistic regression model:

```{r}
model <- glm(PHENOTYPE ~ rs1525873_T, data = geno_subset_dom, family = binomial)
summary(model)
```

Visualize the association between the phenotype and the genotype:

```{r}
# Make prediction over continuous range (for smooth curve)
 geno_range <- data.frame(
   rs1525873_T = seq(0, 1, length.out = 100)
 )
 geno_range$predicted_prob <- predict(model, newdata = geno_range, type = "response")

count_data <- geno_subset_dom %>%
  mutate(PHENOTYPE = round(PHENOTYPE)) %>%
  filter(!is.na(rs1525873_T), !is.na(PHENOTYPE)) %>%
  group_by(rs1525873_T, PHENOTYPE) %>%
  summarise(count = n(), .groups = "drop")

print(count_data)

p2=ggplot() +
  geom_point(
    data = count_data,
    aes(x = rs1525873_T, y = PHENOTYPE, size = count),
    color = "blue"
  ) +
  geom_line(
    data = geno_range,
    aes(x = rs1525873_T, y = predicted_prob),
    color = "red",
    size = 1.2
  ) +
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Logistic Regression: PHENOTYPE ~ Genotype",
    x = "Genotype (0/1)",
    y = "Phenotype ",
    size = "Count"
  ) +
  coord_cartesian(xlim = c(-0.5, 1.5), ylim = c(-0.1, 1.1)) +
  theme_minimal()

print(p2)
#ggsave("./02_activities/data/p2_plot.png", plot = p2, width = 7, height = 5, dpi = 300)
```

**Interpretation**: In the additive model, we see a clear dose-response relationship between genotype and disease risk: each additional copy of the risk allele increases the log-odds of being a case, as reflected in the significant and positive regression coefficient (\~1.93). By contrast, the dominance model treats any presence of the risk allele (Aa or AA) the same, leading to a steeper estimated effect (\~2.76). The AIC is higher under the dominance model (AIC=3705.6), indicating the additive model (AIC=3250.2) may fit slightly better overall by capturing a more gradual allele-dose effect.

The AIC is better when the value is smallest.

##### Recessive Coding:

```{r}
geno_subset_dom <- geno_subset  # make a copy
geno_subset_dom[, 7:ncol(geno_subset_dom)] <- as.data.frame(
  lapply(geno_subset[, 7:ncol(geno_subset)], function(x) ifelse(x == 2, 1, 0))
)
```

```{r}
# Fit logistic regression model
model <- glm(PHENOTYPE ~ rs1525873_T, data = geno_subset_dom, family = binomial)
summary(model)

geno_range <- data.frame(
   rs1525873_T = seq(0, 1, length.out = 100)
 )
 geno_range$predicted_prob <- predict(model, newdata = geno_range, type = "response")

count_data <- geno_subset_dom %>%
  mutate(PHENOTYPE = round(PHENOTYPE)) %>%
  filter(!is.na(rs1525873_T), !is.na(PHENOTYPE)) %>%
  group_by(rs1525873_T, PHENOTYPE) %>%
  summarise(count = n(), .groups = "drop")

print(count_data)

p3=ggplot() +
  geom_point(
    data = count_data,
    aes(x = rs1525873_T, y = PHENOTYPE, size = count),
    color = "blue"
  ) +
  geom_line(
    data = geno_range,
    aes(x = rs1525873_T, y = predicted_prob),
    color = "red",
    size = 1.2
  ) +
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Logistic Regression: PHENOTYPE ~ Genotype",
    x = "Genotype (0/1)",
    y = "Phenotype ",
    size = "Count"
  ) +
  coord_cartesian(xlim = c(-0.5, 1.5), ylim = c(-0.1, 1.1)) +
  theme_minimal()

print(p3)
#ggsave("./02_activities/data/p3_plot.png", plot = p3, width = 7, height = 5, dpi = 300)
```

**Interpretation**: In the recessive model, the predicted risk when genotype = 0 is higher than under the dominance model,when genotype = 0 includes both aa and Aa individuals. This means the average risk for genotype=0 in the recessive model includes some risk from heterozygotes (Aa), pushing the predicted probability higher than if it were only aa. The AIC (AIC=−2log(L)+2k) of 3412.1 indicates that the additive model offers the best overall fit, consistent with the interpretation that heterozygotes (Aa) confer intermediate risk relative to aa.

In practice, the choice of genotype coding depends on biological hypotheses about the mode of inheritance and the statistical fit of different models. Researchers typically start with additive coding as the primary analysis, and may perform sensitivity analyses with dominance and recessive codings.