---
title: "Assignment #2"
format: html
---

You will need **PLINK2** installed and available in your PATH. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). The dataset for this assignment consists of the following binary PLINK files: `gwa.A2.bed`, `gwa.A2.bim`, `gwa.A2.fam` , available at the following Google Drive link: <https://drive.google.com/drive/folders/1rHoy3z52Yyj985ukjjtLhfIBxchpyYtZ?usp=drive_link>. Please download all three files and save them in `02_activities/data/`.

#### Question 1: Data inspection

```{r}
#Set a consistent path for the working directory for the assignment
knitr::opts_knit$set(root.dir = normalizePath("../../"))
```

```{r}
#Need to upload the required libraries
library(data.table)
library(ggplot2)
library(seqminer)
library(dplyr)
library(qqman)
```

Before you run any models, first get familiar with the dataset. You may find `data.table::fread()` in R helpful for reading `.bim` and `.fam` files.

(i) Read the `.fam` file. How many samples does the dataset contain?

```{r}
Fam_file <- fread("./02_activities/data/gwa.A2.fam")
nrow(Fam_file)
head(Fam_file)
cat("The .fam file contains:", nrow(Fam_file), "samples" , "\n")
table(Fam_file$V5)
cat("From the 4000 samples, 2000 are males and 2000 samples are females, based on the number of unique values from column 5 that encodes information about sex", "\n")
```

```{bash}
wc -l ./02_activities/data/gwa.A2.fam
```

The bash command gives the same results, in total, we have 4000 samples in the .fam file

(ii) Read the `.bim` file. How many SNPs does the dataset contain?

```{r}
SNP_Info <- fread("./02_activities/data/gwa.A2.bim")
head(SNP_Info)
nrow(SNP_Info)
length(unique(SNP_Info$V2)) 
cat("The bim file contains:", nrow(SNP_Info), "unique SNPs", "\n")
``` 

```{bash}
head ./02_activities/data/gwa.A2.bim
wc -l ./02_activities/data/gwa.A2.bim
```

In total we have 306,102 SNPs

(iii) Make a histogram (or density plot) of the phenotype. Does it look roughly Gaussian?

```{r}
Pheno_Histo <- ggplot(Fam_file, aes(x = V6)) +
    geom_histogram(fill = "antiquewhite4") +
    labs(title = "Histogram Distribution of Phenotype Severity",
    x = "Phenotype spectrum", y = "Density") + 
    theme_classic()
print(Pheno_Histo)

Pheno_Density <- ggplot(Fam_file, aes(x = V6)) +
    geom_density(fill = "antiquewhite4") +
    labs(title = "Density Distribution of Phenotype severity",
    x = "Phenotype spectrum", y = "Density") +
    theme_classic()
print(Pheno_Density)

cat("Yes, based on the shape of the curve of the distribution of the phenotype, it looks Gaussian. The spectrum of phenotypic values cluster around the mean, and the distribution has a roughly symmetric bell-shaped curve with a single peak at its center, suggesting the data is approximately normally distributed" , "\n")
```

#### Question 2: Quality control (QC)

Now we will perform QC using PLINK2 for the genotype files in `gwa.A2`.

(i) Using PLINK2 from the command line (bash), perform basic QC with the following filters: MAF ≥ 0.05, SNP missingness (`--geno`) ≤ 0.01, individual missingness (`--mind`) ≤ 0.10, and HWE p-value ≥ 0.00005, and output the QC’ed dataset as `gwa.qc.A2`.

```{bash}
plink2 --bfile ./02_activities/data/gwa.A2 --maf 0.05 --geno 0.01 --mind 0.1 --hwe 0.00005 --make-bed --out ./02_activities/data/gwa.qc.A2

wc -l ./02_activities/data/gwa.qc.A2.bed
wc -l ./02_activities/data/gwa.qc.A2.bim
wc -l ./02_activities/data/gwa.qc.A2.fam
```

#### Question 3: Relatedness

In this question, you will use **PLINK2’s built-in KING-robust kinship** (`--king-cutoff`) detect and remove related individuals.

(i)  Perform LD pruning on `gwa.qc.A2` using PLINK2 with the following parameters: `--indep-pairwise 500 50 0.05`, and then generate a new dataset containing only the pruned SNPs.

```{bash}
plink2 --bfile ./02_activities/data/gwa.qc.A2 --indep-pairwise 500 50 0.05 --out ./02_activities/data/gwa.qc.A2
wc -l ./02_activities/data/gwa.qc.A2.prune.in
wc -l ./02_activities/data/gwa.qc.A2.prune.out
```

Then Need to create the new datasets containing the independent SNPs after LD pruning. 
```{bash}
plink2 --bfile ./02_activities/data/gwa.qc.A2 --extract ./02_activities/data/gwa.qc.A2.prune.in --make-bed --out ./02_activities/data/gwa.qc.A2.pruned
wc -l ./02_activities/data/gwa.qc.A2.pruned.bim
wc -l ./02_activities/data/gwa.qc.A2.pruned.fam
wc -l ./02_activities/data/gwa.qc.A2.pruned.bed
```

(ii) Use PLINK2 on the LD-pruned dataset to identify a set of unrelated individuals up to (approximately) 2nd-degree relatives (use a kinship cutoff of 0.0884).

```{bash}
plink2 --bfile ./02_activities/data/gwa.qc.A2.pruned --king-cutoff 0.0884 --out ./02_activities/data/gwa.qc.A2_king
wc -l ./02_activities/data/gwa.qc.A2_king.king.cutoff.in.id
wc -l ./02_activities/data/gwa.qc.A2_king.king.cutoff.out.id
```

(iii) Using the unrelated individual list from part (ii), create a dataset containing only unrelated individuals and name it `gwa.qc_unrelated.A2`.

```{bash}
plink2 --bfile ./02_activities/data/gwa.qc.A2 --keep ./02_activities/data/gwa.qc.A2_king.king.cutoff.in.id --make-bed --out ./02_activities/data/gwa.qc_unrelated.A2
wc -l ./02_activities/data/gwa.qc_unrelated.A2.bed
wc -l ./02_activities/data/gwa.qc_unrelated.A2.bim
wc -l ./02_activities/data/gwa.qc_unrelated.A2.fam
```

#### Question 4: principal component analysis (PCA)

(i) Using PLINK2, perform PCA on the unrelated, LD-pruned dataset to obtain principal components.

    (Hint: Refer to the PCA section in the GWAS Tutorial II. You should use a `*.prune.in` file to select a set of independent SNPs before performing PCA.)

First, It is necessary to create new datasets containing only independent SNPs and unrelated samples 
```{bash}
plink2 --bfile ./02_activities/data/gwa.qc_unrelated.A2 --extract ./02_activities/data/gwa.qc.A2.prune.in --make-bed --out ./02_activities/data/gwa.qc.A2_unrelated_pruned
wc -l ./02_activities/data/gwa.qc.A2_unrelated_pruned.bed
wc -l ./02_activities/data/gwa.qc.A2_unrelated_pruned.bim
wc -l ./02_activities/data/gwa.qc.A2_unrelated_pruned.fam
```

With these new datasets, it is possible to run PCA analysis 
```{bash}
plink2 --bfile ./02_activities/data/gwa.qc.A2_unrelated_pruned --pca 20 --out ./02_activities/data/gwa.qc.A2_unrelated_pruned_PCA
```


#### Question 5: GWAS analyses

We now test for association between each SNP and the continuous phenotype, adjusting for population structure using the top 3 PCs.

Assume that:

-   You will run GWAS on the unrelated, QC’ed dataset `gwa.qc_unrelated.A2`.

-   You have a covariate file (e.g., the `*.eigenvec` output from the PCA analysis) containing PC1–PC3 for each individual.

(i) Using PLINK2, run a linear regression GWAS using `gwa.qc_unrelated.A2` as the input dataset. Adjust for PCs 1–3 as covariates.

```{bash}
plink2 --bfile ./02_activities/data/gwa.qc_unrelated.A2 --ci 0.95 --glm --covar ./02_activities/data/gwa.qc.A2_unrelated_pruned_PCA.eigenvec --covar-name PC1,PC2,PC3 --adjust --out ./02_activities/data/gwa.qc.A2_assoc
```

(ii) Create a Manhattan plot of the GWAS results.

```{r}
A2_assoc_file <- fread("./02_activities/data/gwa.qc.A2_assoc.PHENO1.glm.linear")
head(A2_assoc_file)
nrow(A2_assoc_file)
length(unique(A2_assoc_file$ID))

A2_assoc_Beta_SNP <- A2_assoc_file[TEST == "ADD"]
head(A2_assoc_Beta_SNP)

#Then, it is necessary to change the name of the columns from the PLINK2 output file to meet the specifications used by qqman
setnames(A2_assoc_Beta_SNP, c("#CHROM","POS","ID"), c("CHR","BP","SNP"))

#Then, need to generate the Manhattan Plot
png("./02_activities/data/A2_Manhattan_Plot.png", width = 1000, height = 800)
manhattan(A2_assoc_Beta_SNP,
          chr = "CHR",
          bp  = "BP",
          snp = "SNP",
          p   = "P",
          main = "GWAS Manhattan Plot for Quantitative Phenotype",
          xlab = "Chromosome Number",
          ylab = "-log10 P",
          suggestiveline = FALSE,
          cex.axis = 1.5,
          col = c("lightblue", "lightslateblue"),
          annotatePval=5e-8)
dev.off()
```

(iii) Create a QQ plot of the GWAS p-values.

```{r}
png("./02_activities/data/A2_QQ_Plot.png", width = 800, height = 600)
qq(A2_assoc_Beta_SNP$P, main = "Q-Q Plot of Assignment GWAS p-values")
dev.off()
```

### Criteria

| Criteria | Complete | Incomplete |
|:--------------|:----------------------------------|:----------------------|
| **Data inspection** | Correct sample and SNP counts; phenotype plot with a brief comment on Gaussianity. | Counts or phenotype description/plot missing or incorrect. |
| **QC & LD pruning** | Correct PLINK2 QC command and thresholds. | QC/pruning commands, thresholds, or output datasets missing or incorrect. |
| **Relatedness & PCA** | Correct use of PLINK2 command to obtain unrelated samples and PCA run on pruned SNPs. | Relatedness step, unrelated dataset, or PCA analysis missing or incorrect. |
| **GWAS & visualisation** | Linear regression GWAS with PCs as covariates; Manhattan and QQ plots produced. | GWAS command, or Manhattan/QQ plots missing or clearly incorrect. |

### Submission Information

🚨 **Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md)** 🚨 for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: 11:59 PM – 01/04/2026

-   The branch name for your repo should be: `assignment-2`

-   What to submit for this assignment:
    -   Populate this Quarto document (`assignment_2.qmd`).
    -   Render the document with Quarto: `quarto render assignment_2.qmd`.
    -   Submit both `assignment_2.qmd` and the rendered HTML file `assignment_2.html` in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Create a branch called `assignment-2`.
-   Ensure that your repository is public.
-   Review [the PR description guidelines](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md#guidelines-for-pull-request-descriptions) and adhere to them.
-   Verify that your link is accessible in a private browser window.
-   Confirm that both `assignment_2.qmd` and `assignment_2.html` are included in the pull request.
