---
title: "Assignment #1"
format: html
knitr:
  opts_chunk:
    engine.path: "C:/Program Files/Git/usr/bin/bash.exe"
---

## Assignment 1

You only need to write lines of code for each question. When answering questions that ask you to identify or interpret something, the length of your response doesn’t matter. For example, if the answer is just ‘yes,’ ‘no,’ or a number, you can just give that answer without adding anything else.

We will go through comparable code and concepts in the live learning session. If you run into trouble, start by using the help help() function in R, to get information about the datasets and function in question. The internet is also a great resource when coding (though note that no outside searches are required by the assignment!). If you do incorporate code from the internet, please cite the source within your code (providing a URL is sufficient).

Please bring questions that you cannot work out on your own to office hours, work periods or share with your peers on Slack. We will work with you through the issue.

You will need to install PLINK and run the analyses. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). PLINK is a free, open-source whole genome association analysis toolset, designed to perform a range of basic, large-scale analyses in a computationally efficient manner.

#### Question 1: Data inspection

Before fitting any models, it is essential to understand the data. Use R or bash code to answer the following questions about the `gwa.qc.A1.fam`, `gwa.qc.A1.bim`, and `gwa.qc.A1.bed` files, available at the following Google Drive link: <https://drive.google.com/drive/folders/11meVqGCY5yAyI1fh-fAlMEXQt0VmRGuz?usp=drive_link>. Please download all three files from this link and place them in `02_activities/data/`.

(i) Read the .fam file. How many samples does the dataset contain?

```{r setup, include=FALSE}   
#First need to specify a consistent path for the working directory for the assignment
knitr::opts_knit$set(root.dir = normalizePath("../../"))
```

```{r}
##Then need to upload the libraries to current session of R
library(data.table)
library(ggplot2)
library(seqminer)
library(HardyWeinberg)
library(dplyr)
```

```{r}
Fam_File <- fread("./02_activities/data/gwa.qc.A1.fam")
head(Fam_File)
nrow(Fam_File)
#It contains 4000 samples (since each row represents a different individual)
```

```{bash}
head ./02_activities/data/gwa.qc.A1.fam
wc -l ./02_activities/data/gwa.qc.A1.fam
#This code provides the same results using bash commands
```

(ii) What is the 'variable type' of the response variable (i.e.Continuous or binary)?

```{r}        
head(Fam_File)
length(unique(Fam_File$V6))
#Since the unique possible values for the phenotype variable is the same as the number of samples, this represent a Continuous variable
```

(iii) Read the .bim file. How many SNPs does the dataset contain?

```{r}         
Bim_File <- fread("./02_activities/data/gwa.qc.A1.bim")
head(Bim_File)
nrow(Bim_File)
length(unique(Bim_File$V2))
#It contains 101083 unique SNPs 
```

```{bash}
wc -l ./02_activities/data/gwa.qc.A1.bim   
```

 #### Question 2: Allele Frequency Estimation

(i) Load the genotype matrix for SNPs rs1861, rs3813199, rs3128342, and rs11804831 using additive coding. What are the allele frequencies (AFs) for these four SNPs?

```{bash, results='hide', message=FALSE, warning=FALSE}   
printf "%s\n" rs1861 rs3813199 rs3128342 rs11804831 > ./02_activities/data/SNP_list.A1.txt

plink2 --bfile ./02_activities/data/gwa.qc.A1 --extract ./02_activities/data/SNP_list.A1.txt --recode A --out ./02_activities/data/gwa.qc_Subset_SNPs_additive
head ./02_activities/data/gwa.qc_Subset_SNPs_additive.raw
```

```{r}
Subset_SNPs_Geno_Mtrx <- fread("./02_activities/data/gwa.qc_Subset_SNPs_additive.raw")
head(Subset_SNPs_Geno_Mtrx)

#Compute the allele frequency for each of the four SNPs, ignoring names
Subset_SNPs_allele_freq <- apply(Subset_SNPs_Geno_Mtrx[,7:10], 2, function(x) {
    sum(x, na.rm = T) / (2*sum(!is.na(x)))
})

Subset_SNPs_allele_freq
```

(ii) What are the minor allele frequencies (MAFs) for these four SNPs?

```{bash}         
plink2 --bfile ./02_activities/data/gwa.qc.A1 --extract ./02_activities/data/SNP_list.A1.txt --freq --out ./02_activities/data/gwa.qc_Subset_SNPs_freq
```

```{r}
Subset_SNP_All_Freq <- fread("./02_activities/data/gwa.qc_Subset_SNPs_freq.afreq")
head(Subset_SNP_All_Freq)
cat("The minor allele frequencies (MAFs) for the four SNPs are:", "\n", paste(Subset_SNP_All_Freq$ID, collapse=", "),"are", "\n", paste(Subset_SNP_All_Freq$ALT_FREQS, collapse = " "), "respectively", "\n")

Reference_All_freq <- 1 - Subset_SNP_All_Freq$ALT_FREQS
round(Reference_All_freq, 5) == round(Subset_SNPs_allele_freq, 5)
#This confirms that the results from part I and II are the same
```

#### Question 3: Hardy–Weinberg Equilibrium (HWE) Test

(i) Conduct the Hardy–Weinberg Equilibrium (HWE) test for all SNPs in the .bim file. Then, load the file containing the HWE p-value results and display the first few rows of the resulting data frame.

```{bash}         
plink2 --bfile ./02_activities/data/gwa.qc.A1 --hardy --out ./02_activities/data/gwa.qc.A1_hwe
```

```{r}
HWE_test <- fread("./02_activities/data/gwa.qc.A1_hwe.hardy")
head(HWE_test)
```

(ii) What are the HWE p-values for SNPs rs1861, rs3813199, rs3128342, and rs11804831?

```{r} 
Subset_SNPs <- c("rs3813199", "rs11804831", "rs3128342", "rs1861")  
HWE_p_val <- HWE_test$P[HWE_test$ID %in% Subset_SNPs]
HWE_p_val
cat("The HWE p-values for SNPs:", "\n", paste(Subset_SNPs, collapse = ", "), "are", HWE_p_val, "respectively", "\n")
```

#### Question 4: Genetic Association Test

(i) Conduct a linear regression to test the association between SNP rs1861 and the phenotype. What is the p-value?

```{r}     
#First, need to determine if there is any missing value for the phenotype column or the SNP of interest rs1861 in any of the sample in the Genotype matrix
sum(is.na(Subset_SNPs_Geno_Mtrx$PHENOTYPE))
sum(is.na(Subset_SNPs_Geno_Mtrx$rs1861_C))

#There are 36 samples with missing genotypes for the SNP of interest and these need to be removed.
Subset_SNPs_Geno_Mtrx <- Subset_SNPs_Geno_Mtrx %>% filter(!is.na(Subset_SNPs_Geno_Mtrx$rs1861_C))
head(Subset_SNPs_Geno_Mtrx)
table(Subset_SNPs_Geno_Mtrx$rs1861_C)
Subset_SNP_All_Freq[Subset_SNP_All_Freq$ID == "rs1861",]

model_rs1861 <- glm(PHENOTYPE ~ rs1861_C, data = Subset_SNPs_Geno_Mtrx, family = gaussian)
p_value <- summary(model_rs1861)$coefficient["rs1861_C",4]
p_value
```

(ii) How would you interpret the beta coefficient from this regression?

```{r} 
Beta_rs1861 <- summary(model_rs1861)$coefficient["rs1861_C",1]
Beta_rs1861

cat("The Beta coefficient for SNP rs1861 is:", Beta_rs1861, "\n", "In this case, since we are fitting a linear regression model with a single predictor variable, the beta coefficient represents the slope of the line of best fit. Since the slope is positive and we are using an additive model, this means that for each additional A2 allele present in an individual, the value of the Phenotype increases as a rate equal to the beta coefficient and based on the p value this association is statistically significant", "\n")
```

(iii) Plot the scatterplot of phenotype versus the genotype of SNP rs1861. Add the regression line to the plot.

```{r}         
# Make prediction over continuous range (to draw the line of best fit)
# Create a new data frame containing a sequence of genotype values
geno_range <- data.frame(
   rs1861_C = seq(0, 2, length.out = 100)
 )
# Use the fitted linear regression model to predict the value of the phenotype 
# for each genotype value in geno_range
geno_range$predicted_value <- predict(model_rs1861, newdata = geno_range)

# Start from the observed genotype matrix
count_data <- Subset_SNPs_Geno_Mtrx %>%
# Remove rows with missing genotype or missing phenotype
  filter(!is.na(rs1861_C), !is.na(PHENOTYPE)) %>%
 # Group data by genotype value and phenotype category 
  group_by(rs1861_C, PHENOTYPE) %>%
# Count how many observations fall into each genotype-phenotype combination
  summarise(count = n(), .groups = "drop")

# Start building the plot
A1.p1=ggplot() +
# Add points showing observed genotype-phenotype combinations
  geom_point(
    # Use the summarized count_data for the points
    data = count_data,
    # Map genotype to x-axis, phenotype to y-axis, and count to point size
    aes(x = rs1861_C, y = PHENOTYPE, size = count),
    color = "blue"
  ) +
  # Add a regression line showing predicted values across genotype values
  geom_line(# Use the prediction data frame for the line
    data = geno_range,
    aes(x = rs1861_C, y = predicted_value),
    color = "red",
    # Set line thickness
    size = 1.2
  ) +
  # Control the size range of the points
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Linear Regression Model for the Association of PHENOTYPE ~ rs1861_C Genotype",
    x = "rs1861_C Genotype (0/1/2)",
    y = "Phenotype",
    size = "Count"
  ) +
  # Restrict the visible x- and y-axis range
  coord_cartesian(xlim = c(-0.5, 2.5), ylim = c(-0.1, max(count_data$PHENOTYPE))) +
   # Apply a minimal theme for a clean appearance
  theme_minimal()

print(A1.p1)
```

(iv) Convert the genotype coding for rs1861 to recessive coding.

```{r}      
#Create a new dataset to store the new genotype matrix based on recessive coding    
Recessive_Geno_Mtrx <- Subset_SNPs_Geno_Mtrx  %>% select(FID, IID, PAT, MAT, SEX, PHENOTYPE, rs1861_C)

Recessive_Geno_Mtrx$rs1861_C <- ifelse(Subset_SNPs_Geno_Mtrx$rs1861_C == 2, 1, 0)
table(Subset_SNPs_Geno_Mtrx$rs1861_C)
table(Recessive_Geno_Mtrx$rs1861_C)
#This confirms that the code worked
```

(v) Conduct a linear regression to test the association between the recessive-coded rs1861 and the phenotype. What is the p-value?

```{r}         
Recessive_Model <- glm(PHENOTYPE ~ rs1861_C, data = Recessive_Geno_Mtrx, family = gaussian)
summary(Recessive_Model)
Beta_Rec <- summary(Recessive_Model)$coefficient["rs1861_C", 1]
Rec_p_value <- summary(Recessive_Model)$coefficient["rs1861_C",4]
Rec_p_value
```

(vi) Plot the scatterplot of phenotype versus the recessive-coded genotype of rs1861. Add the regression line to the plot.

```{r}         
# Make prediction over continuous range (to draw the line of best fit)
# Create a new data frame containing a sequence of genotype values
geno_range_rec <- data.frame(
   rs1861_C = seq(0, 1, length.out = 100)
 )
# Use the fitted linear regression model to predict the value of the phenotype 
# for each genotype value in geno_range
geno_range_rec$predicted_value <- predict(Recessive_Model, newdata = geno_range_rec)

# Start from the observed genotype matrix
count_data_rec <- Recessive_Geno_Mtrx %>%
# Remove rows with missing genotype or missing phenotype
  filter(!is.na(rs1861_C), !is.na(PHENOTYPE)) %>%
 # Group data by genotype value and phenotype category 
  group_by(rs1861_C, PHENOTYPE) %>%
# Count how many observations fall into each genotype-phenotype combination
  summarise(count = n(), .groups = "drop")

# Start building the plot
A1.p1_rec = ggplot() +
# Add points showing observed genotype-phenotype combinations
  geom_point(
    # Use the summarized count_data for the points
    data = count_data_rec,
    # Map genotype to x-axis, phenotype to y-axis, and count to point size
    aes(x = rs1861_C, y = PHENOTYPE, size = count),
    color = "blue"
  ) +
  # Add a regression line showing predicted values across genotype values
  geom_line(# Use the prediction data frame for the line
    data = geno_range_rec,
    aes(x = rs1861_C, y = predicted_value),
    color = "red",
    # Set line thickness
    size = 1.2
  ) +
  # Control the size range of the points
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Linear Regression Model for the Association of PHENOTYPE ~ rs1861_C Genotype under Recessive coding",
    x = "rs1861_C Genotype (0/1)",
    y = "Phenotype",
    size = "Count"
  ) +
  # Restrict the visible x- and y-axis range
  coord_cartesian(xlim = c(-0.5, 1.5), ylim = c(-0.1, max(count_data_rec$PHENOTYPE))) +
   # Apply a minimal theme for a clean appearance
  theme_minimal()

print(A1.p1_rec)
```

(vii) Which model fits better? Justify your answer.

```{r}         
summary(model_rs1861)$aic
summary(model_rs1861)$coefficient["rs1861_C", 1]
summary(model_rs1861)$coefficient["rs1861_C", 4]

summary(Recessive_Model)$aic
summary(Recessive_Model)$coefficient["rs1861_C", 1]
summary(Recessive_Model)$coefficient["rs1861_C", 4]

cat("Based on these results, the additive model provides slightly better performance than the recessive model. Although the beta coefficient for the recessive model is slightly higher, meaning that the rate of increase of the value of the phenotype is higher in this model, the p-value is slightly lower compared to the additive model. Furthermore, the AIC of the additive model is slightly lower indicating that the additive model fits slightly better the data", "\n")
```

### Criteria

| Criteria | Complete | Incomplete |
|------------------------|------------------------|------------------------|
| **Data Inspection** | Correct sample/SNP counts and variable type identified. | Missing or incorrect counts or variable type. |
| **Allele Frequency Estimation** | Correct allele and minor allele frequencies computed. | Frequencies missing or wrong. |
| **Hardy–Weinberg Equilibrium Test** | Correct PLINK command and p-value extraction in R. | PLINK command or extraction incorrect/missing. |
| **Genetic Association Test** | Correct regressions, plots, coding, and interpretation. | Regression, plots, or interpretation missing/incomplete. |

### Submission Information

📌 Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md) for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

#### Note:

If you like, you may collaborate with others in the cohort. If you choose to do so, please indicate with whom you have worked with in your pull request by tagging their GitHub username. Separate submissions are required.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: `11:59 PM – 16/03/2026`

-   Branch name for your repo should be: `assignment-1`

-   What to submit for this assignment:

    -   Populate this Quarto document (`assignment_1.qmd`).
    -   Render the document with Quarto: `quarto render assignment_1.qmd`.
    -   Submit both `assignment_1.qmd` and the rendered HTML file `assignment_1.html` in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Created a branch with the correct naming convention.
-   Ensured that the repository is public.
-   Reviewed the PR description guidelines and adhered to them.
-   Verified that the link is accessible in a private browser window.
-   Confirmed that both `assignment_1.qmd` and `assignment_1.html` are included in the pull request.

If you encounter any difficulties or have questions, please don't hesitate to reach out to our team via our Slack help channel. Our technical facilitators and learning support team are here to help you navigate any challenges.
